\thetitle

### 2008-06-13 paper.tex
Applied all of E's changes, except where noted.

\newcommand{\thetitle}{Linear Representation of Network Traffic}
\newcommand{\theauthors}{
  Stefan~Karpinski \and
  Elizabeth~M.~Belding \and
  Kevin~C.~Almeroth \and
  John~R.~Gilbert
}

%\usepackage{type1cm}
%\usepackage[pdftex]{graphicx}
\usepackage[labelfont=bf,small]{caption}
\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array}
\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
\usepackage{mathtools}

\title{\thetitle}
\author{\theauthors}
\institute{
  S.~Karpinski \and
  E.~Belding \and
  K.~Almeroth \and
  J.~Gilbert \\
  Department of Computer Science \\
  University of California, Santa Barbara (USA) \\
  Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
}

@@

\begin{abstract}
We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix. The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations. The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling. We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
%
%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks. Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general. We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally. We conclude with future directions for modeling research based on the new algebraic framework.
\end{abstract}

@@

@@

% In short, the reason is that these assumptions drastically simplify both traffic analysis and workload generation.
% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
% This immediately begs the question of whether these assumptions are benign.
% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.

% TODO: is this paragraph necessary?
% The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
% The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
% Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
% In the terminology introduced above, this is a flow-conditional packet behavior model.
% Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
% In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
% Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
% Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.

If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
There are two reasons.
The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
% This immediately begs the question of whether these assumptions are benign.
% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to understand?
% This paper answers that question.
The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.
% Representing complex multi-level network behaviors in a way that does not make these simplifying assumptions but allows effective traffic analysis and workload generation is currently an unsolved problem;
% we begin to address it here.

% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.

Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
Reduction in representation size and complexity makes analysis and generation feasible.
From this perspective, three questions immediately present themselves:
\begin{enumerate}
\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
\item Are there better ways to simplify and reduce the representation of network behavior?
\end{enumerate}
This paper answers all three questions. We begin, however by answering the second problem and  
using our solution to address the other two questions.
% transforming the unreduced representation of traffic into it's uniform, marginal or conditional counterparts.

@@

To address the the other two questions, we begin with a simple observation.
In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
Moreover, real traffic matrices can be transformed into uniform, marginal and conditional forms through very simple matrix multiplications.
These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
We demonstrate experimentally, that this loss of information is not benign:
uniform and marginal traffic models severely distorting crucial network performance metrics.

@@

While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
This is primarily because wireless networks are locally resource constrained.
Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
If a wired network operator expects larger volumes of traffic on a conditional basis, the solution is to upgrade the hardware, not to change the network protocols.
In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.

Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
He shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.

@@

The behavior of a flow as it affects the network is characterized by the following properties:
its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.);
its source and destination nodes;
its start time;
and the specific sizes packets of and intervals between their transmission.
We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{5em}\=\hspace{4em}\=$\in\Nodes\,$\=\kill
\>\bullet\>types:\>$\type$\>$\in\Types$\>$= \set{1,\dots,255} \subset \N$ \\
\>\bullet\>nodes:\>$\src,\dst$\>$\in\Nodes$\>$= \set{1,\dots,n} \subset \N$ \\
\>\bullet\>times:\>$\start$\>$\in\Times$\>$= \R$ \\
\>\bullet\>sizes:\>$\size$\>$\in\Sizes$\>$= \set{0,\dots,\text{\caps{MTU}}} \subset \N$ \\
\>\bullet\>intervals:\>$\ival$\>$\in\Ivals$\>$ = \R^+ \union \set{\infty}$.
\end{tabbing}
There are 255 possible \caps{IP} types (0 is reserved for \caps{IPv6}).
The number $n$ is simply the total number of nodes in the network being modeled.
\caps{MTU} is the maximum transfer unit of the network (typically around 1500 bytes).
We formalize the packet behavior of a flow as a pair of infinite sequences of packet sizes and inter-packet intervals:
\begin{align}
\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq \\
\ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
\end{align}
Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow. Since flows are finite, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
\begin{align}
\forall\: k: \quad
  \size_k >  \mathrlap{0} \phantom{\infty}
  &\iff k \le \pkts
  \eqn{packet-positivity} \\
\forall\: k: \quad
  \ival_k < \infty
  &\iff k < \pkts.
  \eqn{interval-finiteness}
% \forall\: k: \quad k \le \pkts &\iff \size_k > 0       \eqn{packet-positivity} \\
% \forall\: k: \quad k <   \pkts &\iff \ival_k < \infty. \eqn{interval-finiteness}
\end{align}
Actual packet sizes are required to be positive because we are modeling application layer data transfer:
empty data transmission requests are non-events.
The send time of the $k$th packet can be expressed as
\begin{align}
\time_k = \start + \sum_{i=1}^{k-1} \ival_i.
\end{align}
\Equation{interval-finiteness} implies that this will be finite if and only if $k \le \pkts$.
Finally, we formalize the space of flows:
\begin{align}
  \Flows =
    \Types    \cross
    \Nodes^2  \cross
    \Times    \cross
    \N        \cross
    \SizeSeq  \cross
    \IvalSeq.
\end{align}
We write a generic element of $\Flows$ as
\begin{align}
  \flow = \tuple{
    \type,
    \src,
    \dst,
    \start,
    \pkts,
    \sizes,
    \ivals
  } \in \Flows.
\end{align}
Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
\begin{align}
  \Traffic = \set{X \in 2^\Flows : \card{X} < \infty}.
\end{align}
A traffic sample is simply an element of $\Traffic$. We will refer to elements of $\Traffic$ interchangeably as ``traffic samples'' or ``traffic instances.''

@@

\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly. The simplest is to ignore it; we leave leave more complex schemes to the reader.}

@@

The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
With care, the sum of representations of flows can still naturally describe their aggregate behavior.
Let the representation be $\Phi: \Traffic \to \FlowVec$, where
\begin{align}
\FlowVec = \R^{255} \cross \R^n \cross \R^n \cross L^2[\R] \cross L^2[\R].
\end{align}
% Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
Here $L^2[\R]$ is the vector space of real functions, $\R^\R=\set{f: \R \to \R}$, with the usual norm~\cite{Rudin87}.
% \begin{align}
% \norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
% \end{align}
We define
\begin{align}
\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g},
\end{align}
where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are functions defined by
\begin{align}
f(t) &= \sum_{k=1}^{\pkts}
  \begin{cases}
    1 & \text{if $t \ge \time_k$} \\
    0 & \text{if $t <   \time_k$},
  \end{cases} \\
g(t) &= \sum_{k=1}^{\pkts}
  \begin{cases}
    \size_k & \text{if $t \ge \time_k$} \\
    0       & \text{if $t <   \time_k$}.
  \end{cases}
\end{align}
Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets and total bytes sent, respectively, at time $t$.
From the representation, $\Phi(\flow)$, of a single flow, we can readily recover its type, source, destination, and the transmission times and sizes of packets.

@@

The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins---%
one per value for properties that are discrete already and require no compression---%
and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
In this scheme, representation vectors are effectively histograms, with coordinate values in each dimension counting items falling into the corresponding bin.
% The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
\Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the number of bins, \latin{i.e.} dimension, for the property in question.
% The index assigned determines the histogram bin of the input property.

@@

The final form of homogenization we consider, is conditionality.
This is a restricted form of marginality:
we may assume that marginal distributions apply to all the flows in certain groupings.
For example, we can provide a distribution of flow sizes on a per-source basis.
That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
This is simply accomplished by left-multiplication by a non-uniform matrix.
For example, to make flow behavior uniform per source, consider the ``source-conditioning'' matrix:
$\trans{\mat{S}}\mat{S} \in \R^{n \times n}$.
This matrix has $(\trans{\mat{S}}\mat{S})_{ij} = 1$ if and only if flows $i$ and $j$ have the same source node.
To condition flow behavior by source, we left-multiply by the stochasticized source-condtioning matrix:
\begin{align}
  \mat{F}'= \rStoch(\trans{\mat{S}}\mat{S})\mat{F}.
\end{align}
In this case, rows do not have the same sums, so the ``$\rStoch$'' operator is not equivalent to scalar multiplication.
Per destination uniformity can be achieved similarly, replacing $\mat{S}$ with $\mat{D}$.

@@

The above procedure allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
This allows us to stochastically generate a sequence of packets in accordance with a row from an arbitrary \caps{GMM} matrix instance using the following procedure:
\begin{enumerate}
\item Sample random source and destination node pairs.
Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding collisions.
The simplest technique is the rejection method:
if the same node is chosen for both source and destination, reject that pair and begin the selection process again.
\item Next, randomly sample the number of bytes.
\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
From these, together with the flow size in bytes, we estimate the expected duration of the flow:
$\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
\item Randomly sample a start time for the flow. Ensure that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow by using the rejection method.
\item The first packet is sent at the chosen start time.
The size of each packet is determined by sampling a random size value.
The interval until the next packet is determined by sampling a random interval value.
The packet stream ends when the number of bytes allocated to the flow have been emitted or when the packet emission time exceeds $t_{\max}$.
The last packet is only the size of the remaining number of bytes.
\end{enumerate}
Some remarks about the procedure:
\begin{itemize}
\item For some vectors in $\R^{f \cross N}$, it may be impossible to choose distinct source and destination nodes or flow sizes and start times satisfying the procedure.
However, for any vector that is a sum of vectors representing valid flows---\latin{i.e.} ones having distinct source and destination nodes and feasible combinations of start time, duration, and packet behaviors---the procedure will terminate.
\item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
In the case of a row which is simple a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
\end{itemize}
Despite these remarks, we will show in our experimental evaluation, that this procedure generates \emph{sufficiently realistic} wireless workloads~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.

@@

We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments. We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters.
In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

@@

\input{figures/box_plots}

>>>>

\newcommand{\thetitle}{Linear Representation of Network Traffic}
\newcommand{\thesubtitle}{With Applications to Generating Realistic Wireless Workload}
\newcommand{\theauthors}{
  Stefan~Karpinski \and
  Elizabeth~M.~Belding \and
  Kevin~C.~Almeroth \and
  John~R.~Gilbert
}

\usepackage[labelfont=bf,small]{caption}
\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array,mathtools}
\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle: \thesubtitle}]{hyperref}

\title{\thetitle}
\subtitle{\thesubtitle}
\author{\theauthors}
\institute{
  S.~Karpinski \and
  E.~Belding \and
  K.~Almeroth \and
  J.~Gilbert \\
  Department of Computer Science \\
  University of California, Santa Barbara (USA) \\
  Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
}

@@

\begin{abstract}
We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix. The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations. The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling. We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
\end{abstract}

@@

% TODO: add wireless application into this.

@@

If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
There are two reasons.
The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.

Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
Reduction in representation size and complexity makes analysis and generation feasible.
From this perspective, three questions immediately present themselves:
\begin{enumerate}
\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
\item Are there better ways to simplify and reduce the representation of network behavior?
\end{enumerate}
This paper answers all three questions. We begin, however by answering the second problem and  
using our solution to address the other two questions.

@@

To address the the other two questions, we begin with a simple observation.
In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
Moreover, real traffic matrices can be transformed into uniform, marginal and conditional forms through very simple matrix multiplications.
These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
We demonstrate experimentally that this loss of information is not benign:
uniform and marginal traffic models severely distort crucial network performance metrics.

@@

While our work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
This is primarily because wireless networks are locally resource constrained.
Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
If a wired network operator expects larger volumes of traffic on a conditional basis, the solution is to upgrade the hardware, not to change the network protocols.
In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
The ability to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.

Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
The author shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.

@@

The behavior of a flow as it affects the network is characterized by the following properties:
its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.);
its source and destination nodes;
its start time;
and the specific sizes of packets and intervals between their transmission.
We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{5em}\=\hspace{4em}\=$\in\Nodes\,$\=\kill
\>\bullet\>types:\>$\type$\>$\in\Types$\>$= \set{1,\dots,255} \subset \N$ \\
\>\bullet\>nodes:\>$\src,\dst$\>$\in\Nodes$\>$= \set{1,\dots,n} \subset \N$ \\
\>\bullet\>times:\>$\start$\>$\in\Times$\>$= \R$ \\
\>\bullet\>sizes:\>$\size$\>$\in\Sizes$\>$= \set{0,\dots,\text{\caps{MTU}}} \subset \N$ \\
\>\bullet\>intervals:\>$\ival$\>$\in\Ivals$\>$ = \R^+ \union \set{\infty}$.
\end{tabbing}
There are 255 possible \caps{IP} types (0 is reserved for \caps{IPv6}).
The number $n$ is simply the total number of nodes in the network being modeled.
\caps{MTU} is the maximum transfer unit of the network (typically around 1500 bytes).
We formalize the packet behavior of a flow as a pair of infinite sequences of packet sizes and inter-packet intervals:
\begin{align}
\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq \\
\ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
\end{align}
Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow. Since flows are finite, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
\begin{align}
\forall\: k: \quad
  \size_k >  \mathrlap{0} \phantom{\infty}
  &\iff k \le \pkts
  \eqn{packet-positivity} \\
\forall\: k: \quad
  \ival_k < \infty
  &\iff k < \pkts.
  \eqn{interval-finiteness}
% \forall\: k: \quad k \le \pkts &\iff \size_k > 0       \eqn{packet-positivity} \\
% \forall\: k: \quad k <   \pkts &\iff \ival_k < \infty. \eqn{interval-finiteness}
\end{align}
Actual packet sizes are required to be positive because we are modeling application layer data transfer:
empty data transmission requests are non-events.
The send time of the $k$th packet can be expressed as
\begin{align}
\time_k = \start + \sum_{i=1}^{k-1} \ival_i.
\end{align}
\Equation{interval-finiteness} implies that this will be finite if and only if $k \le \pkts$.
Finally, we formalize the space of flows:
\begin{align}
  \Flows =
    \Types    \cross
    \Nodes^2  \cross
    \Times    \cross
    \N        \cross
    \SizeSeq  \cross
    \IvalSeq.
\end{align}
We write a generic element of $\Flows$ as
\begin{align}
  \flow = \tuple{
    \type,
    \src,
    \dst,
    \start,
    \pkts,
    \sizes,
    \ivals
  } \in \Flows.
\end{align}
Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
\begin{align}
  \Traffic = \set{X \in 2^\Flows : \card{X} < \infty}.
\end{align}
A traffic sample is simply an element of $\Traffic$. We will refer to elements of $\Traffic$ interchangeably as ``traffic samples'' or ``traffic instances.''

@@

\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly. The simplest is to ignore it; we leave more complex schemes to the reader.}

@@

The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
At the other extreme, we can give a ``perfect'' representation of flows, that allows us to completely reconstruct each flow from its representation.
With care, the sum of representations of flows can still naturally describe their aggregate behavior.
Let the representation be $\Phi: \Traffic \to \FlowVec$, where
\begin{align}
\FlowVec = \R^{255} \cross \R^n \cross \R^n \cross L^2[\R] \cross L^2[\R].
\end{align}
% Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
Here $L^2[\R]$ is the vector space of real functions, $\R^\R=\set{f: \R \to \R}$, with the usual norm~\cite{Rudin87}.
% \begin{align}
% \norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
% \end{align}
We define
\begin{align}
\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g},
\end{align}
where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are functions defined by
\begin{align}
f(t) &= \sum_{k=1}^{\pkts}
  \begin{cases}
    1 & \text{if $t \ge \time_k$} \\
    0 & \text{if $t <   \time_k$},
  \end{cases} \\
g(t) &= \sum_{k=1}^{\pkts}
  \begin{cases}
    \size_k & \text{if $t \ge \time_k$} \\
    0       & \text{if $t <   \time_k$}.
  \end{cases}
\end{align}
Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets and total bytes sent, respectively, at time $t$.
From the representation, $\Phi(\flow)$, of a single flow, we can readily recover its type, source, destination, and the transmission times and sizes of packets.

@@

The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins
(one per value for properties that are discrete already and require no compression),
and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
In this scheme, representation vectors are effectively histograms, with coordinate values in each dimension counting items falling into the corresponding bin.
% The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
\Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the number of bins, \latin{i.e.} dimension, for the property in question.
% The index assigned determines the histogram bin of the input property.

@@

The final form of homogenization we consider is conditionality.
This is a restricted form of marginality:
we may assume that marginal distributions apply to all the flows in certain groupings.
For example, we can provide a distribution of flow sizes on a per-source basis.
That is, each source node can have its \textit{own} distribution, and the sizes of its flows are drawn from that distribution, independent of their other properties.
This is simply accomplished by left-multiplication by a non-uniform matrix.
For example, to make flow behavior uniform per source, consider the ``source-conditioning'' matrix:
$\trans{\mat{S}}\mat{S} \in \R^{n \times n}$.
This matrix has $(\trans{\mat{S}}\mat{S})_{ij} = 1$ if and only if flows $i$ and $j$ have the same source node.
To condition flow behavior by source, we left-multiply by the stochasticized source-condtioning matrix:
\begin{align}
  \mat{F}'= \rStoch(\trans{\mat{S}}\mat{S})\mat{F}.
\end{align}
In this case, rows do not have the same sums, so the ``$\rStoch$'' operator is not equivalent to scalar multiplication.
Per destination uniformity can be achieved similarly, replacing $\mat{S}$ with $\mat{D}$.

@@

The above procedure allows us to sample random values for each property in accordance with the behavioral description implied by $\vec{f}$.
This allows us to stochastically generate a sequence of packets in accordance with a row from an arbitrary \caps{GMM} matrix instance using the following procedure:
\begin{enumerate}
\item Sample random source and destination node pairs.
Since a flow must have distinct source and destination nodes, the sampling of endpoints must be done jointly, avoiding collisions.
The simplest technique is the rejection method:
if the same node is chosen for both source and destination, reject that pair and begin the selection process again.
\item Next, randomly sample the number of bytes.
\item The packet size and inter-packet interval vectors, $\vec{z}$ and $\vec{v}$, allow us to compute the expected average packet size and inter-packet intervals for the flow (see~\Section{flow-totals}), call them $\mean{z}$ and $\mean{v}$.
From these, together with the flow size in bytes, we estimate the expected duration of the flow:
$\mean{d} = \fracx{b\mean{v}}{\mean{z}}$.
\item Randomly sample a start time for the flow. Ensure that the start time plus the expected duration, $\mean{d}$, does not exceed the maximum simulation time for the workflow by using the rejection method.
\item The first packet is sent at the chosen start time.
The size of each packet is determined by sampling a random size value.
The interval until the next packet is determined by sampling a random interval value.
The packet stream ends when the number of bytes allocated to the flow have been emitted or when the packet emission time exceeds $t_{\max}$.
The last packet is only the size of the remaining number of bytes.
\end{enumerate}
Some remarks about the procedure:
\begin{itemize}
\item For some vectors in $\R^{f \cross N}$, it may be impossible to choose distinct source and destination nodes or flow sizes and start times satisfying the procedure.
However, for any vector that is a sum of vectors representing valid flows---\latin{i.e.} ones having distinct source and destination nodes and feasible combinations of start time, duration, and packet behaviors---the procedure will terminate.
\item Because the representation $\phi$ models the seven properties independent of each other, it is impossible to do better than to sample the properties independently, modulo the requirements of the procedure.
In the case of a row which is simply a representation of an actual flow, this makes no difference, except for the generation of packets (which are modeled independently).
In the case of composite rows, however, this implies that flow properties are modeled independent of each other.
\end{itemize}
Despite these remarks, we will show in our experimental evaluation, that this procedure generates \emph{sufficiently realistic} wireless workloads~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.

@@

We use the Qualnet wireless network simulator (version 4.0.1) to perform our experiments. We simulate a stationary multi-hop 802.11g network using the Ad hoc On-demand Distance Vector (\caps{AODV}) routing protocol~\cite{rfc:aodv}, with nodes placed randomly in a square field with sides of 150 meters but a radio range of only 20 meters---both in accordance with an indoor network environment.
In addition to the active nodes corresponding to trace \caps{IP}s, half as many passive ``infrastructure'' nodes are added to each simulation: these nodes initiate no data and simply serve as additional network relays. Our simulations resemble multi-hop mesh networks of the kind that are increasingly studied and deployed for delivery of broadband access in residential, corporate and conference settings. We do not attempt to reproduce the physical environment of the original wireless network, nor do we simulate mobility. The only aspect of the original network's behavior that is reproduced is the total pattern of network-wide traffic.

@@

% TODO: remove mixed conditional from plots.
\input{figures/box_plots}
